### 2018-07-06 latex/introduction.tex
First round of including Mathias Z.s comments. Not finished, yet.

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).

We here present a phylogenetic analysis of the NAD pathways in eukaryotes. In contrast to our previous analysis \cite{Gossmann2012FEBS}, which has been limited by the low number of eukaryotic genomes available at the time, we are now able to provide a comprehensive analysis of the eukaryotic pathway evolution. The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the pathway dominant in bacteria is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. To explain these counterintuitive results we built a mathematical model of the pathway and demonstrated that NNMT has a critical role to maintain high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, our model predicts that NNMT likely exerted an evolutionary pressure on the NamPRT affinity development. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete. Based on multiple sequence alignments, we identified a sequence insertion in NamPRT in Deuterostomes. We could show experimentally that this insertion increases the affinity of NamPRT, supporting the predictions derived from our mathematical modelling approach.

>>>>

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. In the first step of the salvage pathway, Nam is converted to the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam is achieved by an extraordinary high affinity of NamPRT to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This extraordinary affinity appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \todo{Needs reference}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. Why is there one enzyme (NamPRT) seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, while at the same time there is another enzyme (NNMT) that seems to have no metabolic function other than to remove Nam from NAD metabolism, This puzzle becomes even more intriguing when considering that the majority of lower organisms and plants deamidate Nam to nicotinic acid (NA) using the nicotinamide deamidase (NADA) before it can enter NAD biosynthesis via the Preiss-Handler pathway (fig.~\ref{fig:pathway_overview}).

We here present a phylogenetic analysis of the NAD pathways. The results suggest that there has been a selection for the co-existence of NamPRT and NNMT in Deuterostomia, while the deamidation pathway, which is dominant in bacteria, is lost. This transition was accompanied by a marked increase in the number of NAD-consuming signalling enzymes. Mathematical modelling of the pathway revealed an unexpected positive kinetic role of NNMT in the maintenance of high NAD-consuming signalling fluxes by preventing accumulation of inhibitory Nam. In addition, the model predicts that NNMT likely exerted an evolutionary pressure on NamPRT to develop a high affinity towards its substrate Nam. Indeed, we identified a short sequence insertion in NamPRT, which first occurs in Deuterostomes that appears to modulate the affinity of NamPRT. Simulating the resource competition we furthermore show that the presence of high affinity NamPRT together with NNMT makes the NADA-dependent pathway obsolete.

### 2018-07-06 latex/results.tex
First round of including Mathias Z.s comments. Not finished, yet.

\subsection{Phylogenetic analysis of NAD biosynthesis and consumption}

@@

Looking across all known species, two different pathways exist that recycle Nam. The major pathway found in yeast and plants uses a four-step pathway starting with the deamination of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amidation step catalysed by NAD synthase, driven by the conversion of ATP to AMP producing pyrophosphate. Even though the latter pathway seems to be very inefficient, it is the pathway preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).

Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail reveals that not only NamPRT replaces NADA, but in most organisms, especially in Deuterostomia, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were not be able to find any gene with considerable similarity in fungi or plants, even though in fungi genes named NNMT can be found in databases. These genes do, however, show very limited sequence similarities to the NNMT of nematodes or deuterostomes. The yeast protein has later been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}. Nematodes are the only organisms that encode NNMT together with NADA without NamPRT being present. In Deuterostomia the only large class that does only have NamPRT and seems to have lost NNMT again are Sauropsida and here especially birds. The reason why a lot of birds do not encode NNMT remains unclear, as the appearance is quite scattered (suppl. fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine. There are some species where we could not find NamPRT or NADA but NNMT, we assume that this is due to incomplete genomes in the database.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we looked at phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so we used the previously established classification into ten different enzyme families \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analysis can be found in supplementary table~S1. The numbers in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxa. With the exception of Cnidaria and Lophotrochozoa, we find average three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

Taken together, we found that the presence of NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling, whilst NADA is lost in vertebrates. This seems counterintuitive, as one would expect that a decrease in precursor concentration caused by the precursor removal through NNMT, should cause a decrease of NAD availability and consequently less active NAD dependent signalling.

\subsection{Dynamics of NAD biosynthesis and consumption}

>>>>

\subsection{Paradoxical evolutionary correlation between NAD-dependent signaling and precursor metabolism}

@@

Two principal pathways exist that recycle Nam. As shown in figure \ref{fig:pathway_overview}, the major pathway found in most prokaryotes, yeast, and plants includes four steps pathway starting with the deamidation of Nam to nicotinic acid by NADA. The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in mammals directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme, NAPRT, converts NA into the NA mononucleotide (NAMN) in the Preiss-Handler pathway. NMN and NAMN are converted into dinucleotides by Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This reaction includes an enzyme adenylation step that consumes ATP. Therefore, even though Nam recycling by NADA is energetically less efficient, this pathway is preferentially used by most bacteria, fungi and plants, whereas most metazoans recycle Nam using NamPRT (fig.~\ref{fig:phylo_distribution}A).

Analysing the phylogeny of the NAD-recycling enzymes in Metazoa in more detail revealed that not only NamPRT replaced NADA, but in most organisms, especially in Deuterostomes, NamPRT is found together with the Nam methyltransferase NNMT (fig.~\ref{fig:phylo_distribution}B). NNMT methylates Nam to methyl-Nam that is in mammals excreted with the urine, thus removing Nam from recycling. NNMT seems to have arisen \textit{de novo} in the common ancestor of Ecdysozoa and Lophotrochozoa. We were unable to find any \textit{NNMT} gene with considerable similarity in fungi or plants. Even though fungal genes named \textit{NNMT} can be found in databases, these genes show very limited sequence similarities to the NNMT of nematodes or deuterostomes. The yeast NNMT protein has meanwhile been shown to be a lysin-protein-methyltransferase \cite{Wlodarski2011}. Nematodes are the only organisms that can have any combination of NADA, NamPRT, and NNMT. In Deuterostomia, the only large clade that does only have NamPRT and seems to have lost NNMT again are Sauropsida and among them especially birds. The reason why a lot of birds do not seem to encode NNMT remains unclear, as the appearance is quite scattered (suppl. fig.~S2). The lack of NNMT might be related to the excretion system, as the product of NNMT methyl-Nam is in mammals excreted with the urine. There are some species for which we could not find NamPRT or NADA, while NNMT was present. We assume that this is due to incomplete genomes in the database.

In addition to the phylogenetic distribution of the two Nam salvage enzymes NADA and NamPRT, we analysed the phylogenetic diversity of enzymes catalysing NAD-dependent signalling reactions. To do so, we used the previously established classification into ten different families of NAD-consuming signalling enzymes \cite{Gossmann2012FEBS}. The detailed list of templates used for the phylogenetic analyses can be found in supplementary table~S1. The numbers shown in figure~\ref{fig:phylo_distribution}B denote the average number of NAD-dependent signalling enzyme families found in each taxon. With the exception of Cnidaria and Lophotrochozoa, we find an average of three to four families in Protostomia, whereas most Deuterostomia have on average more than eight families with an increasing diversification of enzymes within some of these families \cite{Gossmann2014DNAR}.

Taken together, we found that NADA is lost in vertebrates, but strongly preserved in most other organisms, despite its higher energetic requirement. Moreover, the selection for having both NamPRT and NNMT coincides with an increased diversification of NAD-dependent signalling. This observation seems counterintuitive. One would expect that increased NAD-dependent signalling should be accompanied by an increase of substrate availability for NAD biosynthesis. However, NNMT irreversibly removes Nam from salvage back to NAD.

\subsection{Functional properties of NamPRT and NNMT have evolved to maximize NAD-dependent signaling}
